---
title: "Problem Set: Vectors, Strings, Factors, and Dates"
format:
  html:
    toc: true
    toc-depth: 2
    code-overflow: wrap
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(nycflights13)
library(lubridate)
```

------------------------------------------------------------------------

## Question 1 — Exploring the Star Wars Universe

The `dplyr::starwars` dataset (available as soon as you load `tidyverse`) contains physical and biographical information on characters from the Star Wars films. Start by running `View(starwars)` and ?starwars to get oriented.

The code below stores the characters with blue eyes in the "blue_eyes" dataframe, and the characters with red eyes in the "red_eyes" dataframe. There is one character with both blue and red eyes. Use a logical vector approach again with str_detect to find out which one. Hint: one way to do it is with the logical & operator, which returns all rows for which multiple criteria are met

```{r}
# Characters with "blue" in eye_color
blue_eyes <- starwars |>
  filter(str_detect(eye_color, "blue"))
blue_eyes

# Characters with "brown" in eye_color
red_eyes <- starwars |>
  filter(str_detect(eye_color, "red"))
red_eyes
```

## Question 2 — American Unemployment Over Time

The `ggplot2::economics` dataset contains monthly U.S. macroeconomic indicators from 1967 to 2015. The `date` column is already a proper `Date` object.

**(a)** The goal here is to see how the unemployment rate changes over time. Start by plotting the number of unemployed people over time:

```{r}
  ggplot(economics, aes(x = date, y = unemploy)) +
  geom_line()
```

Now use lubridate to plot only the unemployment data from Jan 1st 2008 to Jan 1st 2009. Hint: use the lubridate function ymd to specify date windows, and use a similar logical criteria to in Q1, above

```{r}
library(lubridate)

economics |>
  filter(date >= ymd("2008-01-01") & date <= ymd("2009-01-01")) |>
  ggplot(aes(x = date, y = unemploy)) +
  geom_line()
```

What do you notice about the format of the date variable on the x axis of this plot vs the plot over the entire time window. What do you think is happening here?

**Your answer:** The x-axis in the full dataset shows years spread out over a long time period, so the labels are more spaced and summarized (like showing only years or major intervals). In the filtered 2008–2009 plot, the x-axis shows more detailed date labels (such as months), because the time range is much smaller

This happens because ggplot automatically adjusts the scale of the x-axis depending on the range of the data. When the time window is large, it simplifies the labels to avoid clutter, and when the window is small, it shows more detailed time intervals.

Now use mutate to calculate a new variable - the monthly unemployment rate as `unemploy / pop * 100`. Make a plot over the whole time period as in part 1 of this question.

```{r}
library(dplyr)
library(ggplot2)

economics |>
  mutate(unemp_rate = unemploy / pop * 100) |>
  ggplot(aes(x = date, y = unemp_rate)) +
  geom_line()
```

Do you draw different conclusions now about how unemployment has changed over these 40 years? what explains the difference?

**your answer:** Yes, the conclusion changes. The number of unemployed people increases over time, but the unemployment rate shows more ups and downs rather than a steady rise. This difference is because the population has grown, so the total number can increase even if the percentage unemployed does not.

## Question 3 — Flight Status and the Value of Factors

Run the code below to create a `status` variable that categorises each flight by how early or late it arrived.

```{r}
flight_status <- flights |>
  mutate(
    status = case_when(
      is.na(arr_delay)      ~ "cancelled",
      arr_delay < -30       ~ "very early",
      arr_delay < -15       ~ "early",
      abs(arr_delay) <= 15  ~ "on time",
      arr_delay < 60        ~ "late",
      arr_delay < Inf       ~ "very late",
    ),
    .keep = "used"
  )
```

**(a)** Without converting `status` to a factor, run the followong code to create a bar chart showing the count of flights in each category. What do you notice about the ordering of the bars? Is it meaningful?

```{r}
ggplot(flight_status, aes(x = status)) +
  geom_bar()
```

**your answer:** The bars are ordered alphabetically (cancelled, early, late, etc.), not by how early or late the flights are. This ordering is not meaningful because it does not reflect the natural progression of flight delays.

**(b)** Convert `status` to a factor with levels in a natural severity order: `"very early"`, `"early"`, `"on time"`, `"late"`, `"very late"`, `"cancelled"`. Save it back into `flight_status` and recreate the bar chart. How does the chart change?

```{r}
flight_status <- flight_status |>
  mutate(
    status = factor(
      status,
      levels = c("very early", "early", "on time", "late", "very late", "cancelled")
    )
  )

ggplot(flight_status, aes(x = status)) +
  geom_bar()
```

**your answer:** The chart is now ordered by the severity of delays, from very early to very late, with cancelled flights last. This makes the chart much more meaningful and easier to interpret because it follows a logical progression.
